Classical Planning Algorithms ¡ on the Atari Video Games: Preliminary Results

### 2014-11-18 icaps15-paper/paper.tex
ICAPS-15 paper sources

  
\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}

@@

 \begin{document}
% The file aaai.sty is the style file for AAAI Press 
% proceedings, working notes, and technical reports.
%
\pdfinfo{
/Title (Classical Planning Algorithms on the Atari Video Games: Preliminary Results)
/Author (Nir Lipovetzky, Miquel Ramirez, Hector Geffner)}

@@

\author{Nir, Miquel, H}
%AAAI Press\\
%Association for the Advancement of Artificial Intelligence\\
%2275 East Bayshore Road, Suite 160\\
% Palo Alto, California 94303\\}

@@

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013) 
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however, 
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given, 
and  more importantly,  the action effects and goals are  not known \emph{a priori}. Moreover, in these games there is usually  
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,  
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions, 
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge 
state spaces. The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, 
.., $\IW(k)$ where  $\IW(i)$ is  
a breadth-first search in which a state is  pruned when it is not the first state in the search  to make true some set of i atoms or less.
Preliminary empirical results suggest that the performance of \IW\ with the  $k$ parameter fixed to $1$, 
i.e., $\IW(1)$, is much better than random or breadth-first search, and approaches 
the performance of \UCT. A simple best-first variation  of \IW,  that combine exploration and exploitation,
is evaluated as well. 
\end{quote}
\end{abstract}

@@

The Arcade Learning Environment (ALE)  provides a challenging  platform for evaluating general, domain-independent AI planners and 
learners through a convenient interface to hundreds of Atari 2600 games \cite{bellemare:jair2013}. 
Results have been reported so far for basic planning algorithms like breadth-first 
search and \UCT\ \cite{bellemare:jair2013}, and for  reinforcement learning
and evolutionary  methods \cite{bellemare:jair2013,deep-mind-atari,risto-texas:atari-games}.
The empirical results are impressive in some cases, yet a lot remains to be done, as no method  approaches   the 
performance of  human players across a broad range of games.

While all  these games feature a known initial (RAM) state and actions that
have deterministic effects on the state, the problem of selecting the next action
to be done cannot be addressed with state-of-the-art classical planners \cite{geffner:book}.
This is because there is  no compact PDDL-like encoding of the domain, and more 
importantly,  the goal to be achieved in each game  is not given. Indeed, there are 
often no goals  but rewards $r(a,s)$ that  depend on the action $a$ and 
the state $s$ in  which  the action is performed, and these rewards are not known a priori 
either.\footnote{Actually, in the Atari games, the rewards $r(a,s)$ depend only
on the state $f(a,s)$ that results from doing action $a$ in the state $s$.}
Thus no model of the  goals  or the  rewards can be used to bias the search, and hence, 
standard heuristic search and classical planning methods can't be used.

The action selection problem in the Atari games can be seen as a \emph{reinforcement learning} problem 
\cite{sutton:book} over a deterministic MDP where the state transitions and rewards are not known, or alternatively,
as a  \emph{net-benefit planning problem} \cite{coles2012survey,keyder:jair09} with unknown state transitions  and  rewards. Namely, 
we seek  an action sequence $a_0, a_1, \ldots, a_m$ from the given current  state $s_0$ that generates 
a state sequence $s_0, s_1, \ldots, s_{m+1}$ with  maximum  total reward $\sum_{i=0}^{m} r(a_i,s_i)$,  
where $m$ is a sufficiently large planning horizon, and the rewards $r(a_i,s_i)$ and the  state-transitions $s_{i+1}=f(a_i,s_i)$ 
are known only after  action $a_i$ is applied in the state $s_i$.

@@

The limitation of the   basic blind search  methods is that they are not effective  over large  spaces, neither
for solving problems off-line, nor for guiding  a lookahead search for solving problems on-line. 
In this work, we thus turn to a recent class of  planning algorithms that combine the scope of blind search
methods with the performance of state-of-the-art classical planning algorithms: namely, like ``blind'' search algorithms
they do not require prior knowledge of state transitions, costs, or goals, and yet like heuristic algorithms
they manage to search large  state spaces effectively. The basic algorithm in this class is 
called \IW\ for Iterated Width search \cite{nir:ecai2012}. \IW\ consists of a sequence of calls  $\IW(1)$, $\IW(2)$, .., $\IW(k)$, where 
$\IW(i)$ is  a  standard breadth-first search where states are pruned right away when they fail to make true some new tuple (set) 
of at most $i$ atoms. Namely,   $\IW(1)$ is a breadth-first search that keeps  a state only if  the state is   the first  one 
in the search to make some atom  true; $\IW(2)$ keeps a state  only if the state is  the first one to  make  a pair of atoms true, and 
so on.  Like plain breadth-first  and iterative deepening searches,  \IW\  is 
complete, while searching the state space in a way that makes use of the 
\emph{structure of states} given by the values of a finite set of state variables.
In the Atari games, the (RAM) state is given by a vector of 128 bytes, which we associate with 128 variables 
$X_i$, $i=1, \ldots, 128$, each of which may take up to $256$ values $x_j$. A state $s$ makes an atom $X_i=x_{j}$ 
true when the value of the $i$-th byte in the state vector $s$ is $x_j$.

While a normal, complete breadth-first search runs in time that is exponential in the total number of variables $X_i$, 
the procedure $\IW(k)$ runs in time that is exponential in the $k$ parameter, $1 \leq k \leq n$. 
Elsewhere, it has been shown that  \IW\ exhibits   state-of-the-art performance on most existing planning benchmark domains 
when goals are restricted to be single atoms  where \IW\ provably runs in low polynomial time in the number of
variables \cite{nir:ecai2012}. For example, in a block world instance with any number of blocks and any initial
configuration,  the goal of  placing a given  block on top of another will be achievable in  quadratic time  
in the number of blocks as any such instance can be shown to  have \emph{width} 2. This means  that $\IW(2)$ will be 
not only complete then, but also optimal. This is also true for most other benchmark domains in classical planning
that turn out to  have small, bounded widths for any instance, as long as the goal is atomic. 
For dealing with the benchmarks where goals are  not single atoms, e.g., where a given \emph{tower} of blocks needs to be built, 
simple extensions of the basic \IW\ procedure have been developed as well, such as Serialized \IW, where \IW\ is 
used to achieve the goals one at at time \cite{nir:ecai2012,nir:ecai2014}. 
The extension of \IW\ for problems where there is no goal but an
additive reward measure  to be optimized, like in the Atari games, is direct.

@@

The paper is organized as follows. We review the iterated width  algorithm and where it comes from, 
look at  simple   variations of the algorithm that appear to be convenient  for the Atari
games, present and analyze the preliminary experimental results, and discuss related and future work.

@@

The Iterative Width  (\IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem 
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the 
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to 
represent the value $x_j$ of variable $X_i$ in the state. 
In addition to the state space, a problem is defined by an initial state $s_0$, 
a set of actions applicable in each state, a 
transition function $f$ such that  $s' = f(a,s)$ is 
the state that results from applying action $a \in A(s)$ to the state $s$, 
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not need to be known a priori, yet in that case, the  
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated 
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

\IW consists of a sequence of calls  $\IW(i)$ for $i=0, 1, 2, \ldots$
over a problem $P$ until a termination condition is reached. 
The procedure $\IW(i)$  is a plain forward-state \emph{breadth-first search} with just one change: right after a state 
$s$ is generated, the state is  pruned if it doesn't pass a simple \emph{novelty test}.
More precisely,

\begin{itemize}
\item The \emph{novelty of a newly generate state $s$  in a  breadth-first search}
is   $1$ if  $s$ is the first state generated in the search that makes true some  atom $X=x$, 
else it is $2$ if  $s$ is the first state that makes a \emph{pair} of atoms $X=x$ and $Y=y$ true, 
and so on.

@@

For classical planning, the termination condition is the achievement of the goal. In the on-line setting, 
as in the Atari games, the termination condition is given by a time window or a maximum number of generated nodes.
The  best path found by \IW is the path  that has a  maximum accumulated reward. 
The accumulated reward $R(s)$  of a state $s$ reached in an iteration of \IW is determined by the  unique parent state $s'$ 
and action $a$ leading to $s$ from $s'$ as $R(s) = R(s') + r(a,s')$. The best state is the state $s$ with maximum reward
$R(s)$ generated but not pruned by \IW, and the best path is the one that leads to the state $s$ from the current state.
The action selected to be done next is then the first action along such a path.

@@

\caption{\small 
\emph{Top:} Number of classical planning problems  and percentages solved by $\IW(1)$, $\IW(2)$, or neither. Instances obtained
by splitting benchmarks  with $N$ atomic goals into $N$ problems with atomic goals.
\emph{Bottom:} Number of instances solved by \IW\ in comparison with  Iterative Deepening (\ID),  
Breadth-First Search (\BRFS),   and Greedy Best First Search (\textit{GBFS}) guided by  additive heuristic.
Time and  memory outs  after 30 minutes and 2 GB resp. From \cite{nir:ecai2012}.
}
\label{table1}
\end{table}
%%%%
%%%

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the 
search in a different way. This structured exploration  has proved to be very 
effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition. 
Yet, this changes the domain.}
Table~\ref{table1} from \cite{nir:ecai2012} shows 
the percentage of instances that are  solved by  the first iteration of \IW, i.e.  $\IW(1)$, by the second iteration 
$\IW(2)$, and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows, 
$37\%$ of the 37921 instances are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$. Since $\IW(k)$
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37,921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encoding are all \emph{boolean}.  
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening, 
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuristic $h_{add}$ 
\cite{bonet:aij-hsp} on the other (this algorithm is similar 
to one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}), 
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34,627 vs. 34,849, far ahead
of the other blind algorithms \BRFS\ and \ID\  that solve 9,010 and 8,762 problems each. This is shown in the bottom
part of the Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than the heuristic search  
\cite{nir:ecai2012}.

The min $k$ value for which $\IW(k)$ solves a problem  is clearly bounded and small in most of
these instances. This is  actually no  accident and  has a \emph{theoretical explanation.} Lipovetzky and Geffner define
a structural parameter called the problem \emph{width} and show that for many of these domains,
\emph{any} solvable instance with atomic goals will  have a bounded and small width that is independent of the number
of variables and states in the problem. The min value $k$ for which the iteration  $\IW(k)$ solves
the problem cannot exceed the problem width, so  the algorithm \emph{\IW\ runs in time and space
that are exponential in the problem width.}

@@

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain  
relevant  in  optimization problems as well. Indeed, if a  good path is made of 
states $s_i$ each of which have a low width, \IW\ can be made 
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in \IW\ to enforce this property
which is actually  not yet implemented in the code that is tested below.

@@

The number of nodes \emph{generated} by $\IW(1)$ is $n \times D \times b$ in the worst case, 
where $n$ is the number of variables, $D$ is the size of their domains, and $b$ is the number of 
actions per  state.  This same number in breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by $\IW(2)$ in the worst case is  $(n \times D)^2 \times b$, which is
equal to $1,073,741,824$ which is too large, forcing us to consider only a fraction of 
such states. For classical planning problems, the growth in the number of nodes from 
$\IW(1)$ to $\IW(2)$ is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables, 
and apply these algorithms to that representation. The results would be different.
In such a case, $\IW(1)$ and $\IW(2)$  would generate $n' \times D' \times b$ and 
$(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $7,5497,472$ states respectively.
These are feasible numbers, but we haven't tried  this representation yet
under the assumption that the correlations among the bits in each one of 
the 128 words of the state vectors are meaningful. 
In summary, from the basic \IW\ algorithm we are testing only the first 
call $\IW(1)$.

\IW\ is a purely exploration algorithm that does not take into account  the accumulated 
reward for selecting the next states to consider. As a simple, variant that combines
exploration and exploitation, we evaluated a \emph{best-first search} algorithm
with two queues: one queue  ordered first by  novelty measure 
(recall that novelty one means that the state is the first one to make some atom true),
and a second queue ordered by accumulated reward. In one iteration, the best first
search picks up the best node from one queue, and in the second iteration it picks
up the best node from the other queue. This way for combining multiple heuristics
is used in the LAMA planner \cite{richter:lama}, and was introduced in the
planner Fast Downward \cite{helmert:fd}. In addition, we break ties
in the first queue favoring states with largest accumulated reward, and 
in the second queue, favoring states with smallest novelty measure. Last, when
a node is expanded, it is removed from the queue, and its children are placed on 
both queues. The exception are the nodes with no accumulated reward  that are 
placed in the first queue only. We refer to this best-first algorithm 
as \BFS.

@@

We have shown experimentally that width-based algorithms like $\IW(1)$ and \BFS\ 
that originate in work in classical planning, can be used in the Atari games
with a performance that approaches the state-of-the-art, sometimes using a fraction of the time. We are currently addressing three 
issues, which we haven't managed to get in time for this submission. First, we want to extend the 
evaluation to all the games featured in \cite{bellemare:jair2013}, and to add more runs per  game. 
Second, we want to evaluate the various algorithms for different fixed time windows for decisions.
In the table above, for example, \UCT\ is often taking an order-of-magnitude more time per decision 
than  \BFS. Third, we need to revise the way duplicate states are handled in \IW\ and \BFS\ to ensure
that the best paths among the states that have been expanded are returned. Finally, 
for the future, we also want to explore whether this type of algorithms
can give meaningful results as well  when used on the screen state, as opposed 
to the RAM state.

>>>>

\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}

@@

 \begin{document}
% The file aaai.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%
\pdfinfo{
/Title (Classical Planning Algorithms on the Atari Video Games: Preliminary Results)
/Author ( )}
% (Nir Lipovetzky, Miquel Ramirez, Hector Geffner)}

@@

\author{\textbf{ID:} 52 \ \ ; \ \ \textbf{Keywords:} classical planning, video games}

% \author{Nir Lipovetzky \\
%       The University of Melbourne \\
%    Melbourne, Australia\\
%    {\normalsize\url{@unimelb.edu.au}}
%    \And
%    Miquel Ramirez \\
%    RMIT University \\
%    Melbourne, Australia\\
%   {\normalsize\url{@rmit.edu.au}}
%    \And
%    Hector Geffner \\
%    ICREA \&  Universitat Pompeu Fabra \\
%    Barcelona, SPAIN \\
%    {\normalsize\url{@upf.edu}}\thanks{firstname.lastname}
% }

%AAAI Press\\
%Association for the Advancement of Artificial Intelligence\\
%2275 East Bayshore Road, Suite 160\\
% Palo Alto, California 94303\\}

@@

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013)
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however,
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given,
and  more importantly,  the action effects and goals are  not known \emph{a priori}. Moreover, in these games there is usually
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions,
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge
state spaces. The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, \ldots,
$\IW(k)$ where  $\IW(i)$ is  a breadth-first search in which a state is  pruned when it is not the
first state in the search  to make true some set of $i$ atoms. 
The  empirical results suggest that the performance of \IW\ with the  $k$ parameter fixed to $1$, 
i.e., $\IW(1)$, approaches the performance of the state of the art represented by 
 \UCT, and that a simple best-first variation  of \IW\  that combines exploration 
and exploitation is even more competitive.

\end{quote}
\end{abstract}

@@

The Arcade Learning Environment (ALE)  provides a challenging  platform for evaluating general, domain-independent AI planners and
learners through a convenient interface to hundreds of Atari 2600 games \cite{bellemare:jair2013}.
Results have been reported so far for basic planning algorithms like breadth-first
search and  \UCT,  reinforcement learning algorithms, and evolutionary  methods \cite{bellemare:jair2013,deep-mind-atari,risto-texas:atari-games}.
The empirical results are impressive in some cases, yet a lot remains to be done, as no method  approaches   the
performance of  human players across a broad range of games.

While all  these games feature a known initial (RAM) state and actions that
have deterministic effects, the problem of selecting the next action
to be done cannot be addressed with state-of-the-art classical planners \cite{geffner:book}.
This is because there is  no compact PDDL-like encoding of the domain, and more
importantly,  the goal to be achieved in each game  is not given. Indeed, there are
often no goals  but rewards $r(a,s)$ that  depend on the action $a$ and
the state $s$ in  which  the action is performed, and these rewards are not known.\footnote{Actually, in the Atari games, the rewards $r(a,s)$ depend only
on the state $f(a,s)$ that results from doing action $a$ in the state $s$.}
Thus no model of the  goals  or the  rewards can be used to bias the search.

The action selection problem in the Atari games can be seen as a \emph{reinforcement learning} problem
\cite{sutton:book} over a deterministic MDP where the state transitions and rewards are not known, or alternatively,
as a  \emph{net-benefit planning problem} \cite{coles2012survey,keyder:jair09} with unknown state transitions  and  rewards. 
Namely, we seek  an action sequence $a_0, a_1, \ldots, a_m$ from the  current  state $s_0$ that generates
a state sequence $s_0, s_1, \ldots, s_{m+1}$ with  maximum  total reward $\sum_{i=0}^{m} r(a_i,s_i)$,
where $m$ is a sufficiently large planning horizon, and the rewards $r(a_i,s_i)$ and the  state-transitions $s_{i+1}=f(a_i,s_i)$
are known only after  action $a_i$ is applied in the state $s_i$.

@@

The limitation of the   basic blind search  methods is that they are not effective  over large  spaces, neither
for solving problems off-line, nor for guiding  a lookahead search for solving problems on-line.
In this work, we thus turn to a recent class of  planning algorithms that combine the scope of blind search
methods with the performance of state-of-the-art classical planners: namely, like ``blind'' search algorithms
they do not require prior knowledge of state transitions, costs, or goals, and yet like heuristic algorithms
they manage to search large  state spaces effectively. The basic algorithm in this class is
called \IW\ for Iterated Width search \cite{nir:ecai2012}. \IW\ consists of a sequence of calls  $\IW(1)$, $\IW(2)$, .., $\IW(k)$, where
$\IW(i)$ is  a  standard breadth-first search where states are pruned right away when they fail to make true some new tuple (set)
of at most $i$ atoms. Namely,   $\IW(1)$ is a breadth-first search that keeps  a state only if  the state is   the first  one
in the search to make some atom  true; $\IW(2)$ keeps a state  only if the state is  the first one to  make  a pair of atoms true, and
so on.  Like plain breadth-first  and iterative deepening searches,  \IW\  is
complete, while searching the state space in a way that makes use of the
\emph{structure of states} given by the values of a finite set of state variables.
In the Atari games, the (RAM) state is given by a vector of 128 bytes, which we associate with 128 variables
$X_i$, $i=1, \ldots, 128$, each of which may take up to $256$ values $x_j$. A state $s$ makes an atom $X_i=x_{j}$
true when the value of the $i$-th byte in the state vector $s$ is $x_j$.

While a normal, complete breadth-first search runs in time that is exponential in the total number of variables $X_i$,
the procedure $\IW(k)$ runs in time that is exponential in the $k$ parameter, $1 \leq k \leq n$.
Elsewhere, it has been shown that  \IW\ exhibits   state-of-the-art performance on most existing planning benchmark domains
when goals are restricted to single atoms  where \IW\ provably runs in low polynomial time in the number of
variables \cite{nir:ecai2012}. For example, in a blocks world instance with any number of blocks and any initial
configuration,  the goal of  placing a given  block on top of another will be achievable in  quadratic time
in the number of blocks as any such instance can be shown to  have \emph{width} 2. This means  that $\IW(2)$ will
not only be complete then, but also optimal. This is also true for most other benchmark domains in classical planning
that turn out to  have small, bounded widths for \emph{any} instance as long as the goal is atomic.
For dealing with the benchmarks where goals are  not single atoms, e.g., where a given \emph{tower} of blocks needs to be built,
simple extensions of the basic \IW\ procedure have been developed  such as Serialized \IW, where \IW\ is
used to achieve the goals one at at time following a simple goal ordering  \cite{nir:ecai2012,nir:ecai2014}.
The extension of \IW\ for problems where there is no goal but an
additive reward measure  to be optimized, like in the Atari games, is direct.

@@

The paper is organized as follows. We review the iterated width  algorithm and where it comes from,
look at  simple   variations of the algorithm that appear to be convenient  for the Atari
games, present and analyze the  experimental results, and discuss related and future work.

@@

The Iterated Width  (\IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means of  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to
represent the value $x_j$ of variable $X_i$ in the state.
In addition to the state space, a problem is defined by an initial state $s_0$,
a set of actions applicable in each state, a
transition function $f$ such that  $s' = f(a,s)$ is
the state that results from applying action $a$  to the state $s$,
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not need to be known a priori, yet in that case, the
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

\IW \ consists of a sequence of calls  $\IW(i)$ for $i=0, 1, 2, \ldots$
over a problem $P$ until a termination condition is reached.
The procedure $\IW(i)$  is a plain forward-state \emph{breadth-first search} with just one change: right after a state
$s$ is generated, the state is  pruned if it doesn't pass a simple \emph{novelty test}.
More precisely,

\begin{itemize}
\item The \emph{novelty of a newly generate state $s$  in a  search algorithm}
is   $1$ if  $s$ is the first state generated in the search that makes true some  atom $X=x$,
else it is $2$ if  $s$ is the first state that makes a \emph{pair} of atoms $X=x$ and $Y=y$ true,
and so on.

@@

For classical planning, the termination condition is the achievement of the goal. In the on-line setting,
as in the Atari games, the termination condition is given by a time window or a maximum number of generated nodes.
The  best path found by \IW \ is the path  that has a  maximum accumulated reward.
The accumulated reward $R(s)$  of a state $s$ reached in an iteration of \IW \ is determined by the  unique parent state $s'$
and action $a$ leading to $s$ from $s'$ as $R(s) = R(s') + r(a,s')$. The best state is the state $s$ with maximum reward
$R(s)$ generated but not pruned by \IW, and the best path is the one that leads to the state $s$ from the current state.
The action selected to be done next is  the first action along  such a path.

@@

\caption{\small
\emph{Top:} Number of classical planning instances per domain  and percentages solved by $\IW(1)$, $\IW(2)$, or neither.
Problems  obtained by splitting benchmarks  with $N$ atomic goals into $N$ problems with atomic goals.
\emph{Bottom:} Number of instances solved by \IW\ in comparison with  Iterative Deepening (\ID),
Breadth-First Search (\BRFS),   and Greedy Best First Search (\textit{GBFS}) guided by  additive heuristic.
Time and  memory outs  after 30 minutes or 2 GB. Table from \cite{nir:ecai2012}.
}
\label{table1}
\end{table}
%%%%
%%%

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the
search in a different way. This structured exploration  has proved to be very
effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition.
Yet, this changes the definition of the domain.}
Table~\ref{table1} from \cite{nir:ecai2012} shows
the percentage of instances that are  solved by  the first iteration of \IW, i.e.  $\IW(1)$, by the second iteration
$\IW(2)$, and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows,
$37\%$ of the 37921 instances are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$. Since $\IW(k)$
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37,921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encoding are all \emph{boolean}.
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening,
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuristic $h_{add}$
\cite{bonet:aij-hsp} on the other (this algorithm is similar
to the one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}),
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34,627 vs. 34,849, far ahead
of the other blind algorithms \BRFS\ and \ID\  that solve 9,010 and 8,762 problems each. This is shown in the bottom
part of  Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than a heuristic search algotrithm
\cite{nir:ecai2012}.

The min $k$ value for which $\IW(k)$ solves a problem  is bounded and small in most of
these instances. This is  actually no  accident and  has a \emph{theoretical explanation.} Lipovetzky and Geffner define
a structural parameter called the problem \emph{width} and show that for many of these domains,
\emph{any} solvable instance with atomic goals will  have a bounded and small width that is independent of the number
of variables and states in the problem. The min value $k$ for which the iteration  $\IW(k)$ solves
the problem cannot exceed the problem width, so  the algorithm \emph{\IW\ runs in time and space
that are exponential in the problem width.}

@@

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain
relevant  in  optimization problems as well. Indeed, if a  good path is made of
states $s_i$ each of which has a low width, \IW\ can be made
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in \IW\ to enforce this property.
% which is actually  not yet implemented in the code that is tested below.

@@

The number of nodes \emph{generated} by $\IW(1)$ is $n \times D \times b$ in the worst case,
where $n$ is the number of variables, $D$ is the size of their domains, and $b$ is the number of
actions per  state.  This same number in breadth-first search is not linear in $n$
but exponential. For the Atari games,  $n=128$, $D=256$, and $b=18$, so that the product is
equal to $589,824$, which is large but feasible. On the other hand, the number
of nodes generated by $\IW(2)$ in the worst case is  $(n \times D)^2 \times b$, which is
equal to $1,073,741,824$ which is too large, forcing us to consider only a fraction of
such states. For classical planning problems, the growth in the number of nodes from
$\IW(1)$ to $\IW(2)$ is not that large, as the variables are boolean. Indeed, we could have
taken the state vector for the Atari games as a vector of 1024 boolean variables,
and apply these algorithms to that representation. The results would be different.
In such a case, $\IW(1)$ and $\IW(2)$  would generate $n' \times D' \times b$ and
$(n' \times D')^2 \times b$ states, which for $n'=1024$ and $D=2$
represent  $36,864$ and $7,5497,472$ states respectively.
These are feasible numbers, but we haven't used  this representation 
under the assumption that the correlations among the bits in each one of
the 128 words of the state vectors are meaningful.
In summary, from the basic \IW\ algorithm we are testing only the first
call $\IW(1)$.

\IW\ is a purely exploration algorithm that does not take into account  the accumulated
reward for selecting the states to consider. As a simple variant that combines
exploration and exploitation, we evaluated a \emph{best-first search} algorithm
with two queues: one queue  ordered first by  novelty measure
(recall that novelty one means that the state is the first one to make some atom true),
and a second queue ordered by accumulated reward. In one iteration, the best first
search picks up the best node from one queue, and in the second iteration it picks
up the best node from the other queue. This way for combining multiple heuristics
is used in the LAMA planner \cite{richter:lama}, and was introduced in the
planner Fast Downward \cite{helmert:fd}. In addition, we break ties
in the first queue favoring states with largest accumulated reward, and
in the second queue, favoring states with smallest novelty measure. Last, when
a node is expanded, it is removed from the queue, and its children are placed on
both queues. The exception are the nodes with no accumulated reward  that are
placed in the first queue only. We refer to this best-first algorithm
as \BFS.

@@

We have shown experimentally that width-based algorithms like $\IW(1)$ and \BFS\
that originate in work in classical planning,  can be used to play the Atari video
games and to achieve  state-of-the-art performance. This is the result of a 
structured exploration of the state that manages to combine the scope of blind
search algorithms with the efficiency of heuristic search methods. 
%
\Omit{
We are currently addressing three
issues, which we haven't managed to get in time for this submission. First, we want to extend the
evaluation to all the games featured in \cite{bellemare:jair2013}, and to add more runs per  game.
Second, we want to evaluate the various algorithms for different fixed time windows for decisions.
In the table above, for example, \UCT\ is often taking an order-of-magnitude more time per decision
than  \BFS. Third, we need to revise the way duplicate states are handled in \IW\ and \BFS\ to ensure
that the best paths among the states that have been expanded are returned.Finally,}
%
There are many issues worth studying, we just mention two of them.
One problem of all planning methods in Atari  games is that by using the 
RAM state of the Atari console they ``cheat'' a bit, as they use information that
is not available to the human player. In addition, a limitation of our
algorithms is that they do not get better with experience.  
In the future, we would like to explore whether  variations of these algorithms 
can be used to play  from the information in the  screen state as opposed to the RAM state, 
and the potential uses of the  ideas underlying these algorithms for learning.
